### 2018-07-13 latex/discussion.tex
Incorporated all changes and comments by MZ. Comments not directly addressed are included as todos.

NAD-consuming enzymes are involved in a wide variety of signalling and gene regulatory mechanisms that, due to their sensitivity to NAD$^{+}$, have the ability to translate differences in metabolic states into changes in signalling and gene regulation. As NAD concentrations are lowered by the removal of NAD precursor by NNMT, a high-affinity enzyme for recycling of Nam is required to ensure maintenance for cellular NAD levels for other metabolic processes. It is unclear, why NamPRT developed a higher affinity and not NADA. NADA may have thermodynamic or mechanistic limitations, other environmental conditions may apply, or it might just have happened by chance. Whatever the reason, we show in our simulations that a high affinity of NamPRT is required for high NAD consumption fluxes and NAD concentrations in the presence of NNMT. It therefore seems plausible that NNMT might have been driving NamPRT evolution. Looking at the enzyme affinities of the human enzymes, it furthermore appears that both NNMT and NamPRT reached an almost optimal state, as further changes in the affinity of either NamPRT or NNMT would not result in much higher steady state NAD concentrations or NAD consumption fluxes. In addition, our simulations suggest that NNMT makes both NAD concentration and NAD consumption relatively independent of other processes requiring NAD, such as cell growth.

@@

Neither the scattered distribution of NamPRT and NADA that is especially pronounced in bacteria \cite{Gazzaniga2009} but that has also been observed in eukaryotes, nor the disappearance of NADA in vertebrates has not been understood earlier. Our combined phylogenetic-modelling analysis now provides a potential explanation for both observations. Using simulated competition between two compartments that share for the same limited source of Nam, we show that the compartment that contains NamPRT and NNMT has a higher steady state NAD concentration and NAD consumption rate than the compartment containing NADA. This is, however, only the case if NamPRT substrate affinity is sufficiently high. The dominant enzyme combination found in vertebrates, a high-affinity NamPRT with NNMT, thus seems to provide a competitive advantage. As this may also hold for mammalian-associated bacteria, particularly pathogens, we wanted to see whether pathogenic bacteria solely express NamPRT. Unfortunately, bacterial habitat information is currently far from complete and often difficult to access. We therefore manually checked bacteria that possess NamPRT and indeed found that most of them have been characterised to be pathogenic. It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree \cite{Gazzaniga2009}. Besides the suggestion made here, there might might well be other environmental aspects that influence the phylogenetic distribution in bacteria.

@@

In our analysis, we completely neglected the potential effects of co-substrates of the investigated pathway. Neglected co-substrates include targets of the NAD-consuming enzymes, such as acylated proteins for sirtuins, for example, or phosphoribosyl pyrophosphate (PRPP) and ATP that are required for NMN synthesis by NamPRT. Furthermore, the presence of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT \cite{Ulanovskaya2013} was not considered here. As co-substrate availability might alter the behaviour of the system, these should be included in future analyses. Information about the \textit{in~vivo} concentration of those co-substrates is, however, currently very limited. We anticipate improvements as more sensitive methods for metabolite measurements are currently developed \cite{Frederick2017} \todo{Mathias Z., do you know others?}.

>>>>

\todo[inline]{MZ: Ist Geschmackssache. Ich bin ja mehr so der Typ: "Three major observations have emerged from the present study. 1., 2., 3..." Aber es geht natürlich auch so wie es steht.}

NAD-consuming enzymes are involved in a wide variety of signalling and gene regulatory mechanisms that, due to their sensitivity to NAD$^{+}$, have the ability to translate differences in metabolic states into changes in signalling and gene regulation. As NAD concentrations are lowered by the removal of NAD precursor by NNMT, a high-affinity enzyme for recycling of Nam is required to ensure maintenance for cellular NAD levels for other metabolic processes. It is unclear, why NamPRT developed a higher affinity and not NADA. NADA may have thermodynamic or mechanistic limitations, other environmental conditions may apply, or it might just have happened by chance \todo{MZ: Irgendwo vorher steht, dass dieser Weg energetisch ungünstiger ist. Wäre das nicht zumindest eine überlegenswerte Erklärung?}. Whatever the reason, we show in our simulations that a high affinity of NamPRT is required for high NAD consumption fluxes and NAD concentrations in the presence of NNMT. It therefore seems plausible that NNMT might have been driving NamPRT evolution. Looking at the enzyme affinities of the human enzymes, it furthermore appears that both NNMT and NamPRT reached an almost optimal state, as further changes in the affinity of either NamPRT or NNMT would not result in much higher steady state NAD concentrations or NAD consumption fluxes. In addition, our simulations suggest that NNMT makes both NAD concentration and NAD consumption relatively independent of other processes requiring NAD, such as cell growth.

@@

Neither the scattered distribution of NamPRT and NADA that is especially pronounced in bacteria \cite{Gazzaniga2009} but that has also been observed in eukaryotes, nor the disappearance of NADA in vertebrates has not been understood earlier. Our combined phylogenetic-modelling analysis now provides a potential explanation for both observations. Using simulated competition between two compartments that share for the same limited source of Nam, we show that the compartment that contains NamPRT and NNMT has a higher steady state NAD concentration and NAD consumption rate than the compartment containing NADA. This is, however, only the case if NamPRT substrate affinity is sufficiently high. The dominant enzyme combination found in vertebrates, a high-affinity NamPRT with NNMT, thus seems to provide a competitive advantage. As this may also hold for mammalian-associated bacteria, particularly pathogens, we wanted to see whether pathogenic bacteria solely express NamPRT. Unfortunately, bacterial habitat information is currently far from complete and often difficult to access. We therefore manually checked bacteria that possess NamPRT and indeed found that most of them have been characterised to be pathogenic. It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree \cite{Gazzaniga2009}. Besides the suggestion made here, there might well be other environmental aspects that influence the phylogenetic distribution in bacteria.

@@

In our analyses, we did not consider the potential effects of co-substrates of the investigated pathway. Such co-substrates include targets of the NAD-consuming enzymes, such as acylated proteins for sirtuins, for example, or phosphoribosyl pyrophosphate (PRPP) and ATP that are required for NMN synthesis by NamPRT. Furthermore, the presence of the methyl donor \textit{S}-adenosyl methionine (SAM) and its precursor methionine that have been shown to potentially limit the effect of NNMT \cite{Ulanovskaya2013} was not considered here. As co-substrate availability might alter the behaviour of the system, these should be included in future analyses. Unfortunately, information about the \textit{in~vivo} concentrations of these co-substrates is currently very limited.

### 2018-07-13 latex/figurelegends.tex
Incorporated all changes and comments by MZ. Comments not directly addressed are included as todos.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig_5_optimal_substr_affinities}
  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} We simulated the impact of changes in the $K_{M}$ for both NamPRT and NNMT on NAD consumption rates (A) and NAD concentration (B). Both are increasing with decreasing $K_{M}$ of NamPRT but increasing $K_{M}$ of NNMT. The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to optimal, as further improvements would have little or no effect on NAD consumption or concentration.}
  \label{fig:optimal_substr_affinities}
\end{figure}

>>>>

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig_5_optimal_substr_affinities}
  \caption{\textbf{The substrate affinities of human NNMT and NamPRT are optimal.} We simulated the impact of changes in the $K_{M}$ for both NamPRT and NNMT on NAD consumption rates (A) and NAD concentration (B). Both are increasing with decreasing $K_{M}$ of NamPRT but increasing $K_{M}$ of NNMT. The affinities reported for human enzymes (indicated by a black asterisk) appear to be close to the theoretical optimum, as further improvements would have little or no effect on NAD consumption or concentration.}
  \label{fig:optimal_substr_affinities}
\end{figure}

### 2018-07-13 latex/header.tex
Incorporated all changes and comments by MZ. Comments not directly addressed are included as todos.

\noindent
\textsuperscript{2}Department of Molecular Biology, University of Bergen, Thormøhlensgt. 55, 5020 Bergen, Norway

@@

NAD is best known as cofactor in redox reactions, but it is also substrate of NAD-dependent signalling reactions that consume NAD and release nicotinamide (Nam). In eukaryotes, two different Nam salvage pathways exist. Phylogenetic analysis of the pathway show that in lower organisms the initial deamidation of Nam is prevalent, whilst the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates. Strikingly, loss of the deamidation pathway in early vertebrates is preceded by the occurrence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).  Despite the occurrence of this Nam degrading enzyme, we observe a diversification of NAD dependent signalling enzymes in vertebrates. This seems counterintuitive.

To better understand these evolutionary changes and the role of NNMT for NAD metabolism, we built a mathematical model of the pathway using published enzyme kinetics data. Our results indicate that NNMT is required to enable high NAD consumption fluxes enabling diversification of the NAD-dependent signalling pathways. This kinetic regulation requires a high substrate affinity of the key enzyme for Nam salvage, NamPRT. Indeed, the affinity of NamPRT to Nam has previously been measured to be in the nanomolar range. Based on our mathematical modelling approach we furthermore suggest that NNMT exerted an evolutionary pressure on NamPRT, enforcing the development of its unusually high substrate affinity. Using multiple sequence alignments, we identified a sequence insertion, first occurring in vertebrates, that parallels an experimentally verified increase in the substrate affinity of the enzyme. Additional simulations show that the deamidation pathway became obsolete owing to the high substrate affinity of NamPRT. Collectively, our results illustrate a close evolutionary relationship between NAD biosynthesis and the diversification NAD-dependent signalling pathways, potentially driven by the concomitant occurrence of a regulator of Nam salvage, NNMT.

>>>>

\noindent
\textsuperscript{2}Department of Biomedicine, University of Bergen, Jonas Lies Vei 91, 5020 Bergen, Norway

@@

NAD is best known as cofactor in redox reactions, but it is also substrate of NAD-dependent signalling reactions that consume NAD and release nicotinamide (Nam). Two different Nam salvage pathways exist. We conducted extensive phylogenetic analyses of these pathways and show that in lower organisms the initial deamidation of Nam is prevalent, whilst the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates. Strikingly, loss of the deamidation pathway in early vertebrates is preceded \todo{Statt "preceded", schlägt MZ "accompanied" for. Ich (MB) denke, dass preceded richtig ist.} by the emergence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT). Paradoxically, the occurrence of this Nam degrading enzyme is paralleled by a diversification of NAD dependent signalling enzymes in vertebrates.

To better understand these evolutionary changes and the role of NNMT for NAD metabolism, we built a mathematical model of the pathway using available enzyme kinetics data. Our simulations indicate that NNMT is required to maintain high NAD consumption fluxes, thereby enabling diversification of the NAD-dependent signalling pathways. This kinetic regulation requires and explains the unusually high substrate affinity of the key enzyme for Nam salvage, NamPRT. Moreover, we suggest that NNMT exerted evolutionary pressure on NamPRT, enforcing the development of its high substrate affinity. Using multiple sequence alignments, we identified a sequence insertion, first occurring in vertebrates, that parallels an experimentally verified increase in the substrate affinity of the enzyme. Additional simulations show that the deamidation pathway became obsolete owing to the high substrate affinity of NamPRT. Collectively, our results illustrate a close evolutionary relationship between NAD biosynthesis and the diversification of NAD-dependent signalling pathways, potentially driven by the concomitant occurrence of a regulator of Nam salvage, NNMT.

### 2018-07-13 latex/introduction.tex
Incorporated all changes and comments by MZ. Comments not directly addressed are included as todos.

NAD metabolism represents one of the most critical links that connect cellular signal transduction and energy metabolism. Even though it is best known as cofactor for various redox-reactions, NAD is involved in a number of signalling processes that consume NAD$^{+}$ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}. These NAD-dependent signalling reactions include but are not limited to poly- and mono-ADP-ribosylation \cite{Butepage2015,DeVos2012}, NAD-dependent protein deacylation by sirtuins \cite{Osborne2016}, and the synthesis of calcium-mobilizing molecules such as cyclic ADP-ribose \cite{Lee2012}. These NAD-dependent signalling processes participate in the regulation of virtually all cellular activities. The enzymes involved in these processes are sensitive to the available NAD concentration \cite{Ruggieri2015}, which in turn is dependent on the NAD$^{+}$/NADH redox ratio. Therefore, NAD-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.

@@

Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD synthesis pathway. If Nam would not be constantly recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}. Two principal pathways exist that recycle Nam. Firstly, vertebrates use a direct two step pathway starting with the conversion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate. The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high substrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}. This appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core \cite{Burgos2009}. Despite the importance of its salvage, Nam can also be marked for excretion by methylation. The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism. While NamPRT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no metabolic function other than to remove Nam from NAD metabolism. It has been suggested that the process potentially acts as a metabolic methylation sink \cite{Pissios2017}.

Secondly, in many prokayotes as well as in plants and fungi, a pathway consisting of four steps starting with the deamidation of Nam to nicotinic acid (NA) by the Nam deamidase (NADA) is used. (\ref{fig:pathway_overview}). The three enzymes that act after NADA belong to the Preiss-Handler pathway that also exists in vertebrates. NA is converted into the corresponding mononucleotide (NAMN), in a reactions performed by the NA-specific phosphoribosyltransferase NAPRT. The conversion of the mononucleotides NMN and NAMN into their dinucleotides NAD and NAAD is performed by the same enzyme, the Nam/NA adenylytransferases (NMNATs) that is essential in all organisms \cite{DeFigueiredo2011}. The recycling pathway via NA finally requires re-amidation of the dinucleotide, which is catalysed by NAD synthase. This final reaction includes an enzyme adenylation step that consumes ATP. Therefore the Nam recycling by NADA is energetically less efficient under normal conditions than the recycling pathway starting with NamPRT.

>>>>

NAD metabolism represents one of the most critical links that connect cellular signal transduction and energy metabolism. Even though it is best known as cofactor for various redox-reactions, NAD is involved in a number of signalling processes that consume NAD$^{+}$ by cleaving the molecule to nicotinamide (Nam) and ADP-ribose \cite{Verdin2015}. These NAD-dependent signalling reactions include poly- and mono-ADP-ribosylation \cite{Butepage2015,DeVos2012}, NAD-dependent protein deacylation by sirtuins \cite{Osborne2016}, and the synthesis of calcium-mobilizing molecules such as cyclic ADP-ribose \cite{Lee2012}. These NAD-dependent signalling processes participate in the regulation of virtually all cellular activities. The enzymes involved in these processes are sensitive to the available NAD concentration \cite{Ruggieri2015}, which in turn is dependent on the NAD$^{+}$/NADH redox ratio. Therefore, NAD-dependent signalling can act as a transmitter of changes in the cellular energy homeostasis, for example, to regulate gene expression or metabolic activity \cite{Koch-Nolte2009}.

@@

Due to the constant release of Nam through NAD-consuming signalling reactions, the NAD salvage pathway using Nam as precursor is the most important NAD synthesis pathway. If Nam were not continuously recycled into NAD, humans would require a much higher daily vitamin B3 intake than the 16\,mg that are the current daily recommendation \cite{CommissionofEuropeanCommunities2008}. Two principal pathways exist that recycle Nam. Firstly, vertebrates use a direct two step pathway starting with the conversion of Nam into the mononucleotide NMN by the Nam phosphoribosyltransferase (NamPRT) using phosphoribosyl pyrophosphate (PRPP) as co-substrate. The nearly complete recycling of Nam by NamPRT is achieved by an extraordinary high substrate affinity to Nam, the $K_{M}$ being in the low nanomolar range \cite{Burgos2008}. This appears to be mediated by an ATP-dependent phosphorylation of a histidine residue in the catalytic core \cite{Burgos2009}. Despite the importance of its salvage, Nam can also be marked for excretion by methylation. The presence of nicotinamide N-methyltransferase (NNMT) in vertebrates \cite{Gossmann2012FEBS} is among the most enigmatic and counterintuitive features of NAD metabolism. While NamPRT is seemingly optimised to recycle even the faintest amounts of Nam back into NAD synthesis, NNMT seems to have no metabolic function other than to remove Nam from NAD metabolism. It has been suggested that the process potentially acts as a metabolic methylation sink \cite{Pissios2017}.

Secondly, in most prokayotes as well as in plants and fungi, a pathway consisting of four steps starting with the deamidation of Nam to nicotinic acid (NA) by the Nam deamidase (NADA) is used. (fig.~\ref{fig:pathway_overview}). The three enzymes that act after NADA belong to the Preiss-Handler pathway that also exists in vertebrates. NA is converted into the corresponding mononucleotide (NAMN), in a reactions performed by the NA-specific phosphoribosyltransferase NAPRT. The conversion of both mononucleotides, NMN and NAMN, into their corresponding dinucleotides, NAD and NAAD, is catalysed by the Nam/NA adenylytransferases (NMNATs) that are essential in all organisms \cite{DeFigueiredo2011}. The recycling pathway via NA finally requires re-amidation of NAAD by NAD synthase. This final reaction includes an enzyme adenylation step that consumes ATP. Therefore, the Nam recycling by NADA is energetically less efficient under normal conditions than the recycling pathway starting with NamPRT. \todo{MZ: Hier müssten Fragen/Probleme aufgeworfen werden, z.B.: Wieso sind die beiden Möglichkeiten offenbar evolutionär getrennt? Warum gibt es ein weiteres Enzym, NNMT, das Nam "abbaut"? ...}
